---
title: "Lecture 3: Hypothesis Testing, Power & Comparative Tests"
subtitle: "SISSA PhD in Neuroscience - Practical Lab"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
filters:
  - shinylive
---

## Step 1: Intuition Playground (30 min)

### The Power, Effect Size, and Sample Size Trade-Off
In null-hypothesis significance testing (NHST), we specify:
* Null Hypothesis ($H_0$): $\mu_1 - \mu_2 = 0$
* Significance Threshold ($\alpha$): Type I error rate (false positive)
* Statistical Power ($1 - \beta$): Probability of correctly rejecting $H_0$ given a true effect size $\delta = (\mu_1 - \mu_2)/\sigma$ (Cohen's $d$)

Use the interactive power calculator below:
1. Increase the sample size $n$ per group and notice how the sampling distribution of the alternative hypothesis pulls away from the critical rejection boundary.
2. In neuroscience experiments, small animal cohorts ($n = 4$ to $6$) often suffer from low power ($< 40\%$). What effect size is required to achieve standard $80\%$ power with $n = 5$?
3. Toggle the $\alpha$ slider from $0.05$ to $0.01$: observe the direct penalty imposed on power.

```{shinylive-r}
#| standalone: true
#| viewerHeight: 580

library(shiny)

ui <- fluidPage(
  titlePanel("Statistical Power & Type I / Type II Errors"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("n_per_group", "Sample Size per Group (n):", min = 3, max = 50, value = 10, step = 1),
      sliderInput("cohen_d", "True Effect Size (Cohen's d):", min = 0.1, max = 2.0, value = 0.8, step = 0.05),
      sliderInput("alpha_lvl", "Significance Level (α):", min = 0.01, max = 0.10, value = 0.05, step = 0.01),
      helpText("Shaded green indicates statistical power (1 - β). Shaded red on the null curve indicates α (false positive).")
    ),
    mainPanel(
      plotOutput("powerPlot", height = "460px"),
      verbatimTextOutput("powerSummary")
    )
  )
)

server <- function(input, output) {
  p_calc <- reactive({
    n <- input$n_per_group
    d <- input$cohen_d
    alpha <- input$alpha_lvl
    
    # Standard error of difference in means (assuming pooled sd = 1)
    se <- sqrt(2 / n)
    crit_val <- qnorm(1 - alpha / 2) * se
    
    # Power = P(Z > crit_val | mean = d) + P(Z < -crit_val | mean = d)
    power <- 1 - pnorm(crit_val, mean = d, sd = se) + pnorm(-crit_val, mean = d, sd = se)
    list(se = se, crit = crit_val, power = power, d = d, alpha = alpha)
  })

  output$powerPlot <- renderPlot({
    res <- p_calc()
    x <- seq(-3, res$d + 4, length.out = 600)
    
    y_null <- dnorm(x, mean = 0, sd = res$se)
    y_alt  <- dnorm(x, mean = res$d, sd = res$se)
    
    par(mar = c(4.5, 4.5, 3, 1))
    plot(x, y_null, type = "l", lwd = 2.5, col = "black",
         ylim = c(0, max(c(y_null, y_alt)) * 1.15),
         xlab = "Difference Between Means", ylab = "Density",
         main = paste0("Power (1 - β): ", round(res$power * 100, 1), "% | Sample Size n = ", input$n_per_group))
    lines(x, y_alt, lwd = 2.5, col = "forestgreen", lty = 2)
    
    # Fill Type I Error (alpha)
    x_rejection <- x[x >= res$crit]
    polygon(c(res$crit, x_rejection, max(x_rejection)),
            c(0, dnorm(x_rejection, 0, res$se), 0),
            col = rgb(1, 0, 0, 0.35), border = NA)
            
    # Fill Power (1 - beta)
    polygon(c(res$crit, x_rejection, max(x_rejection)),
            c(0, dnorm(x_rejection, res$d, res$se), 0),
            col = rgb(0.1, 0.7, 0.2, 0.4), border = NA)
            
    abline(v = res$crit, col = "firebrick", lwd = 2, lty = 3)
    legend("topright", legend = c("H0: No Effect (d = 0)", "H1: True Effect (d)", "Rejection Threshold"),
           col = c("black", "forestgreen", "firebrick"), lty = c(1, 2, 3), lwd = 2, bty = "n")
  })
  
  output$powerSummary <- renderPrint({
    res <- p_calc()
    cat("Critical Difference Threshold:", round(res$crit, 3), "\n")
    cat("Probability of Type II Error (Beta):", round(1 - res$power, 4), "\n")
  })
}

shinyApp(ui, server)
```

---

## Step 2: "Code-Along & Fix-It" Lab (85 min)

We load the hippocampal slice dataset: `data/lecture3_synaptic_plasticity.csv`.

### Exercise 1: Paired vs. Two-Sample Independent $t$-Test

In our experiment, each hippocampal slice has a baseline fEPSP slope measured *before* theta-burst stimulation and a post-stimulation response measured *in the same slice*.

```{r}
ltp_data <- read.csv("../data/lecture3_synaptic_plasticity.csv")
ctrl_slices <- subset(ltp_data, group == "Control")

head(ctrl_slices[, c("slice_id", "baseline_fepsp", "post_fepsp")])
```

#### The Paired Design Advantage:
Because baseline responses vary from slice to slice (biological variability), treating repeated measures as independent samples inflates the residual error variance and weakens statistical sensitivity.

Let's verify this mathematically by running both tests:

```r
# --- TASK: COMPARE PAIRED VS UNPAIRED DESIGNS ---
# 1. Independent two-sample test (INCORRECT DESIGN FOR THIS EXPERIMENT)
t_unpaired <- t.test(ctrl_slices$post_fepsp, ctrl_slices$baseline_fepsp, paired = FALSE)

# 2. Paired t-test (ACCOUNTING FOR WITHIN-SLICE REPEATED MEASURES)
t_paired   <- t.test(ctrl_slices$post_fepsp, ctrl_slices$baseline_fepsp, paired = TRUE)

cat("Unpaired p-value:", format.pval(t_unpaired$p.value, digits = 3), "\n")
cat("Paired p-value:  ", format.pval(t_paired$p.value, digits = 3), "\n")
```

Observe how pairing eliminates slice-to-slice baseline shifts, yielding a much tighter test statistic and higher sensitivity.

---

### Exercise 2: The Student's vs. Welch's $t$-Test Trap

Student's classic two-sample $t$-test assumes equal population variances ($\sigma_1^2 = \sigma_2^2$). In neurobiology, pharmacological or optogenetic perturbations frequently alter both the mean and the variance of neural responses.

Compare the LTP magnitude (`post_fepsp`) between `Control` and `Drug_Washout`:

```r
# --- BROKEN CODE CHALLENGE ---
# The researcher runs a classical Student's t-test with var.equal = TRUE:
broken_test <- t.test(post_fepsp ~ group, data = ltp_data, var.equal = TRUE)
print(broken_test)

# Check the variance difference:
tapply(ltp_data$post_fepsp, ltp_data$group, var)
```

#### The Fix:
When sample sizes or group variances differ, Student's pooled variance estimate is biased. **Welch's $t$-test** adjusts the degrees of freedom using the Satterthwaite approximation and does not assume equal variances:

```r
# FIX: Use Welch's t-test (default in R)
welch_test <- t.test(post_fepsp ~ group, data = ltp_data, var.equal = FALSE)
print(welch_test)
```

---

### Exercise 3: Non-Parametric Alternatives (Wilcoxon / Mann-Whitney)

Count data such as spontaneous miniature EPSC frequencies (`mepsc_freq_hz`) or action potential counts are strictly non-negative, discrete, and strongly right-skewed.

```{r}
# Inspect skewness and normality
par(mfrow = c(1, 2))
hist(ltp_data$mepsc_freq_hz[ltp_data$group == "Control"],
     col = "steelblue", main = "Control mEPSCs", xlab = "Hz")
hist(ltp_data$mepsc_freq_hz[ltp_data$group == "Drug_Washout"],
     col = "salmon", main = "Drug mEPSCs", xlab = "Hz")
```

Because Gaussian assumptions are violated, we apply the two-sample non-parametric Wilcoxon Rank-Sum test (Mann-Whitney $U$ test):

```r
# Non-parametric rank-based comparison
wilcox_result <- wilcox.test(mepsc_freq_hz ~ group, data = ltp_data, exact = FALSE)
print(wilcox_result)
```

---

### Exercise 4: Categorical Contingency Analysis ($\chi^2$ & Fisher's Exact)

We recorded whether slices achieved synaptic extinction learning (`1` = Success, `0` = Failure).

```r
# Build 2x2 contingency table
tab <- table(ltp_data$group, ltp_data$extinction_success)
colnames(tab) <- c("Failure", "Success")
print(tab)

# Chi-Square Test of Independence
chi_res <- chisq.test(tab, correct = TRUE) # Yates continuity correction
print(chi_res)

# When cell counts are small (n < 5 in any cell), use Fisher's Exact Test:
fisher_res <- fisher.test(tab)
print(fisher_res)
```

---

## Step 3: Peer Code Review & Diagnostic Discussion (20 min)

With your lab partner, evaluate these three methodological diagnostic questions:

1. **Defaulting to Welch:** Why is it good scientific practice to make Welch's $t$-test (`var.equal = FALSE`) your default choice over Student's $t$-test, even if variances look similar?
2. **Effect Size vs. $p$-value:** If an experiment reports $p = 0.0001$ with $n = 1000$ per group, does that guarantee the biological effect is meaningful? Calculate Cohen's $d$ using `lsr::cohensD()` or manual computation:
   $$d = \frac{\bar{x}_1 - \bar{x}_2}{s_{\text{pooled}}}$$
3. **Paired Permutation Intuition:** How does a paired test fundamentally reduce to a single-sample test on differences: $D_i = X_{\text{post}, i} - X_{\text{pre}, i}$ tested against $H_0: \mu_D = 0$?