---
title: "Lecture 1: Exploratory Data Analysis, Covariance & PCA"
subtitle: "SISSA PhD in Neuroscience - Practical Lab"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
filters:
  - shinylive
---

## Step 1: Intuition Playground (30 min)

### The Geometry of Dimensionality Reduction
In the theoretical lecture, we discussed how covariance measures shared variance between variables and how PCA finds orthogonal axes that maximize projected variance[cite: 73, 75].

Interact with the sandbox below:
1. Adjust the **Projection Vector Angle** $\theta$.
2. Observe the right panel: how does the variance of the projected data points change?
3. Locate the angle that maximizes the projected variance (this corresponds to the first eigenvector, or **PC1**)[cite: 75].

```{shinylive-r}
#| standalone: true
#| viewerHeight: 520

library(shiny)

ui <- fluidPage(
  titlePanel("Maximizing Projected Variance"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("theta", "Projection Vector Angle (°):", min = 0, max = 180, value = 20, step = 2),
      helpText("The red line shows the chosen unit vector. The histogram displays the distribution of data points projected onto this line.")
    ),
    mainPanel(
      plotOutput("pcaPlot", height = "400px")
    )
  )
)

server <- function(input, output) {
  set.seed(123)
  x_raw <- rnorm(70, mean = 0, sd = 1.8)
  y_raw <- 0.7 * x_raw + rnorm(70, mean = 0, sd = 0.7)
  X <- cbind(x_raw - mean(x_raw), y_raw - mean(y_raw))

  output$pcaPlot <- renderPlot({
    rad <- input$theta * pi / 180
    u <- c(cos(rad), sin(rad))
    projections <- as.vector(X %*% u)
    var_proj <- var(projections)

    par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1))

    # Scatter plot with projection line
    plot(X[, 1], X[, 2], pch = 19, col = rgb(0.1, 0.4, 0.8, 0.7),
         xlim = c(-5, 5), ylim = c(-5, 5), asp = 1,
         xlab = "Neuron 1 (centered)", ylab = "Neuron 2 (centered)",
         main = "Original Data Space")
    abline(0, tan(rad), col = "firebrick", lwd = 2.5)
    arrows(0, 0, 3 * cos(rad), 3 * sin(rad), col = "firebrick", lwd = 3, length = 0.1)
    grid()

    # Histogram of projections
    hist(projections, breaks = seq(-6, 6, by = 0.8), col = "darkorange", border = "white",
         xlim = c(-6, 6), ylim = c(0, 35),
         xlab = "Projected Coordinates",
         main = paste("Projected Variance:", round(var_proj, 3)))
    abline(v = mean(projections), col = "black", lty = 2, lwd = 2)
  })
}

shinyApp(ui, server)
```

---

## Step 2: "Code-Along & Fix-It" Lab (80 min)

We will analyze multi-neuron 2-photon calcium imaging recordings: `data/lecture1_calcium.csv`.

### Exercise 1: Loading Data and Summary Statistics

Load the dataset and inspect the measures of central tendency and dispersion introduced in lecture: mean, median, variance, and quantiles[cite: 51, 53, 58, 70].

```{r}
# 1. Load the data
calcium <- read.csv("../data/lecture1_calcium.csv")
head(calcium)

# 2. Compare mean and median on a noisy ROI (roi_2)
mean_roi2 <- mean(calcium$roi_2)
median_roi2 <- median(calcium$roi_2)

cat("Mean ROI 2:   ", round(mean_roi2, 3), "\n")
cat("Median ROI 2: ", round(median_roi2, 3), "\n")
```

#### Trap 1: Outliers and Systematic vs. Unsystematic Error
Notice the discrepancy between mean and median in `roi_2`[cite: 51, 53]. Isolated artifacts (such as preparation movement) distort the sample mean while leaving the median stable[cite: 51, 53, 64].

Identify outliers using the boxplot whisker rule from lecture[cite: 64]:
$$W_l = Q_1 - 1.5 \cdot \text{IQR}, \quad W_u = Q_3 + 1.5 \cdot \text{IQR}$$

```r
# --- TASK: COMPLETE THE CODE ---
# Calculate Q1, Q3, and IQR for calcium$roi_2
q1 <- quantile(calcium$roi_2, 0.25)
q3 <- quantile(calcium$roi_2, 0.75)
iqr_val <- q3 - q1

# Define upper threshold
upper_limit <- q3 + 1.5 * iqr_val

# Extract trials that exceed this threshold:
# FIX ME: complete indexing
outliers <- calcium[calcium$roi_2 > upper_limit, c("trial_id", "condition", "roi_2")]
print(outliers)
```

---

### Exercise 2: Manual Covariance Matrix Construction

The sample covariance between variables $X$ and $Y$ is defined as[cite: 73]:
$$s_{x,y} = \frac{1}{n-1}\sum_{i=1}^n (x_i - \bar{x})(y_i - \bar{y})$$

For a column-centered matrix $X_c$ ($n \times p$), the covariance matrix is given by[cite: 74]:
$$C = \frac{1}{n-1} X_c^T X_c$$

Verify this matrix identity in R:

```{r}
# Subset neurons roi_1 through roi_6
rois <- calcium[, paste0("roi_", 1:6)]
n <- nrow(rois)

# Center columns by subtracting each sample mean
X_c <- scale(rois, center = TRUE, scale = FALSE)

# Compute covariance via matrix multiplication (%*%)
cov_manual <- (1 / (n - 1)) * (t(X_c) %*% X_c)

# Compare with base R cov()
cov_r <- cov(rois)

# Confirm maximum discrepancy is negligible
max(abs(cov_manual - cov_r))
```

---

### Exercise 3: The Scaling Trap in PCA

Run Principal Component Analysis using `prcomp()`.  
In our dataset:
* `roi_7` has very low amplitude values ($\mu \approx 0.05$).
* `roi_8` has a large amplifier baseline offset ($\mu \approx 450, \sigma \approx 65$).

#### The Broken Code:
Observe what happens when PCA is run without unit-variance scaling (`scale. = FALSE`):

```{r}
# Include all 8 recording channels
rois_all <- calcium[, paste0("roi_", 1:8)]

# BUG: scaling disabled
pca_unscaled <- prcomp(rois_all, scale. = FALSE)

# Check variance explained
summary(pca_unscaled)
```

**PC1 accounts for >99.8% of the total variance.**

Inspect the loadings (eigenvector weights) for PC1:
```{r}
pca_unscaled$rotation[, 1]
```

`roi_8` dominates the entire analysis due strictly to its unit scale, masking the biological dynamics of the remaining neurons[cite: 75].

#### The Fix:
Correct the command by setting `scale. = TRUE`:

```r
# --- TASK: COMPLETE THE CODE ---
# Re-run PCA with each channel standardized to unit variance:
pca_scaled <- prcomp(rois_all, scale. = TRUE) # FIX HERE

summary(pca_scaled)
```

---

### Exercise 4: Visualizing Behavioral Trials in Latent Space

Plot the trials in the reduced coordinate system (**PC1 vs. PC2**) to evaluate whether population activity separates `Stimulus` and `Spontaneous` conditions without prior labels[cite: 75].

```r
# Extract coordinates (scores) along PC1 and PC2
scores_df <- as.data.frame(pca_scaled$x)
scores_df$condition <- calcium$condition

# Scatter plot of low-dimensional projections
plot(scores_df$PC1, scores_df$PC2,
     col = ifelse(scores_df$condition == "Stimulus", "firebrick", "steelblue"),
     pch = 19, cex = 1.3,
     xlab = "PC1", ylab = "PC2",
     main = "Trial Projections in PCA Space")
legend("topright", legend = c("Stimulus", "Spontaneous"),
       col = c("firebrick", "steelblue"), pch = 19)
grid()
```

---

## Step 3: Peer Code Review & Diagnostic Discussion (25 min)

Form pairs and compare outputs:

1. **Biological Interpretation of Loadings:**  
   Run `pca_scaled$rotation[, 1:2]`. Which neurons contribute most to PC1? Do the neurons separating *Stimulus* from *Spontaneous* correspond to the sensory-responsive units?
2. **Variance vs. Causality:**  
   Review the caveat from lecture: PCA maximizes variance, not correlation with experimental conditions[cite: 75]. If an artifact (e.g., optical drift or animal motion) introduces large variance, PCA will isolate the artifact as PC1[cite: 75].