---
title: "Lecture 2: Continuous Distributions, CLT & Confidence Intervals"
subtitle: "SISSA PhD in Neuroscience - Practical Lab"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
filters:
  - shinylive
---

## Step 1: Intuition Playground (30 min)

### The "Dance of Confidence Intervals" Sandbox
In the theoretical lecture, we saw that the sample mean $\bar{X}$ is itself a random variable with sampling variance $\sigma^2 / n$. Its confidence interval is given by:
$$IC_{1-\alpha} = \left[\bar{x} - t_{n-1, 1-\frac{\alpha}{2}}\frac{s}{\sqrt{n}}, \; \bar{x} + t_{n-1, 1-\frac{\alpha}{2}}\frac{s}{\sqrt{n}}\right]$$

A frequent mistake is thinking that a 95% confidence interval means *"there is a 95% probability that the true parameter $\mu$ lies within this specific interval."* 

Use the simulator below to simulate 50 repeated experiments from a population with true mean $\mu = 30$:
1. Check how many calculated intervals actually intersect the true dashed line $\mu = 30$.
2. Change the sample size $n$ from 5 to 50: observe what happens to interval widths.
3. Compare the coverage between the normal z-quantile and the Student's t-quantile when sample size $n$ is very small ($n = 4$).

```{shinylive-r}
#| standalone: true
#| viewerHeight: 560

library(shiny)

ui <- fluidPage(
  titlePanel("Sampling Distribution & Confidence Intervals"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("n_samples", "Sample Size (n per experiment):", min = 3, max = 60, value = 8, step = 1),
      sliderInput("conf_level", "Confidence Level (1 - α):", min = 0.80, max = 0.99, value = 0.95, step = 0.01),
      radioButtons("dist_type", "Interval Method:",
                   choices = c("Student's t-quantile (Exact)" = "t", "Gaussian z-quantile (CLT approx)" = "z")),
      actionButton("resim", "Re-run 50 Experiments", class = "btn-primary")
    ),
    mainPanel(
      plotOutput("ciPlot", height = "460px")
    )
  )
)

server <- function(input, output) {
  sim_data <- reactive({
    input$resim
    mu <- 30
    sigma <- 8
    k <- 50
    n <- input$n_samples
    
    means <- numeric(k)
    lower <- numeric(k)
    upper <- numeric(k)
    
    alpha <- 1 - input$conf_level
    
    for (i in 1:k) {
      samp <- rnorm(n, mean = mu, sd = sigma)
      m <- mean(samp)
      s <- sd(samp)
      se <- s / sqrt(n)
      
      crit <- if (input$dist_type == "t") qt(1 - alpha/2, df = n - 1) else qnorm(1 - alpha/2)
      
      means[i] <- m
      lower[i] <- m - crit * se
      upper[i] <- m + crit * se
    }
    
    covers <- (lower <= mu) & (upper >= mu)
    list(means = means, lower = lower, upper = upper, covers = covers, mu = mu)
  })

  output$ciPlot <- renderPlot({
    res <- sim_data()
    k <- length(res$means)
    coverage_rate <- mean(res$covers) * 100
    
    par(mar = c(4.5, 4.5, 3, 1))
    plot(0, 0, type = "n", xlim = c(15, 45), ylim = c(1, k),
         xlab = "Estimated Mean (with Confidence Interval)", ylab = "Experiment Index (1 to 50)",
         main = paste0("Empirical Coverage: ", coverage_rate, "% (Nominal: ", round((1 - (1 - input$conf_level))*100), "%)"))
    
    abline(v = res$mu, col = "black", lwd = 2, lty = 2)
    
    for (i in 1:k) {
      col_line <- if (res$covers[i]) "steelblue" else "firebrick"
      lines(c(res$lower[i], res$upper[i]), c(i, i), col = col_line, lwd = 1.8)
      points(res$means[i], i, pch = 20, col = col_line, cex = 0.9)
    }
    legend("topright", legend = c("Captures True μ", "Misses True μ", "True μ (Ground Truth)"),
           col = c("steelblue", "firebrick", "black"), lty = c(1, 1, 2), lwd = 2, bty = "n")
  })
}

shinyApp(ui, server)
```

---

## Step 2: "Code-Along & Fix-It" Lab (85 min)

We use single-unit timing data: `data/lecture2_neural_timing.csv`.

### Exercise 1: Empirical CDF vs. Theoretical Distributions

In continuous distributions, the probability of obtaining an exact value is zero ($P(X = y) = 0$). We evaluate intervals using the Cumulative Distribution Function (CDF):
$$F_X(t) = P(X \le t) = \int_{-\infty}^t f_X(x)\,dx$$

Let's test whether baseline resting membrane potential (`v_rest_mv`) matches a Gaussian model $\mathcal{N}(\mu, \sigma^2)$.

```{r}
# 1. Load the dataset
timing <- read.csv("../data/lecture2_neural_timing.csv")
v_rest <- timing$v_rest_mv

# 2. Compute empirical summary estimates
mu_hat <- mean(v_rest)
s_hat <- sd(v_rest)

# 3. Compute Empirical Cumulative Distribution Function (ECDF)
ecdf_vrest <- ecdf(v_rest)

# Plot empirical CDF against theoretical Gaussian CDF
grid_vals <- seq(min(v_rest) - 2, max(v_rest) + 2, length.out = 200)
plot(ecdf_vrest, verticals = TRUE, do.points = FALSE, col = "steelblue", lwd = 2,
     main = "Empirical vs. Theoretical Normal CDF",
     xlab = "Membrane Potential (mV)", ylab = "P(V <= t)")
lines(grid_vals, pnorm(grid_vals, mean = mu_hat, sd = s_hat), col = "firebrick", lwd = 2, lty = 2)
legend("topleft", legend = c("Empirical ECDF", "Theoretical Gaussian"),
       col = c("steelblue", "firebrick"), lty = c(1, 2), lwd = 2, bty = "n")
```

---

### Exercise 2: Building Quantile-Quantile (Q-Q) Plots Manually

From the theoretical lecture, a Q-Q plot compares empirical sample quantiles $Q(q)$ against theoretical quantiles $Q_P(q) = F_X^{-1}(q)$ for probability steps $q \in (0, 1)$.

Below is an intentionally broken function that attempts to build a normal Q-Q plot for inter-spike intervals (`isi_ms`):

```r
# --- BROKEN CODE TASK ---
# The function below has two bugs:
# 1. Probabilities q are wrongly spaced or out of range.
# 2. It compares against theoretical quantiles of the wrong distribution without parameter estimates.

broken_qq_plot <- function(data_vec) {
  n <- length(data_vec)
  sorted_data <- sort(data_vec)
  
  # BUG 1: Probabilities must be within (0, 1). Using 1:n includes 1.0 where qnorm() returns +Inf
  # FIX: Use the standard Hazen or Weibull grid: q = (1:n - 0.5) / n
  q_grid <- (1:n) / n # <-- FIX THIS LINE
  
  # Sample quantiles
  sample_quantiles <- sorted_data
  
  # BUG 2: Theoretical quantiles need to be standardized using sample mean and sd
  theoretical_quantiles <- qnorm(q_grid, mean = 0, sd = 1) # Standard normal
  
  plot(theoretical_quantiles, sample_quantiles,
       xlab = "Theoretical Normal Quantiles",
       ylab = "Empirical Quantiles",
       main = "Normal Q-Q Plot", pch = 19, col = "purple")
  abline(lm(sample_quantiles ~ theoretical_quantiles), col = "red", lwd = 2)
}

# Run the function on ISI data
broken_qq_plot(timing$isi_ms)
```

**Task:** Fix line 11 so that `q_grid <- (1:n - 0.5) / n`. Observe the severe right-tail curvature. Why does inter-spike interval data deviate systematically from the straight line?

---

### Exercise 3: Simulating the Central Limit Theorem (CLT)

The CLT states that the standardized mean of independent, identically distributed variables converges to $\mathcal{N}(0, 1)$ as $n \to \infty$, regardless of the underlying distribution (as long as $\sigma^2 < \infty$).

Let's draw samples from the skewed `isi_ms` distribution with sample sizes $n = 3, 10, 50$:

```{r}
set.seed(42)
n_reps <- 2000

# Function to generate sampling distribution of the mean
sample_means <- function(pop, n, reps) {
  replicate(reps, mean(sample(pop, size = n, replace = TRUE)))
}

means_n3  <- sample_means(timing$isi_ms, n = 3,  reps = n_reps)
means_n10 <- sample_means(timing$isi_ms, n = 10, reps = n_reps)
means_n50 <- sample_means(timing$isi_ms, n = 50, reps = n_reps)

par(mfrow = c(1, 3))
hist(means_n3,  breaks = 30, col = "salmon", border = "white", main = "n = 3 (Skewed)", xlab = "Mean ISI")
hist(means_n10, breaks = 30, col = "orange", border = "white", main = "n = 10", xlab = "Mean ISI")
hist(means_n50, breaks = 30, col = "forestgreen", border = "white", main = "n = 50 (Gaussian)", xlab = "Mean ISI")
```

---

### Exercise 4: Standard Deviation vs. Standard Error vs. Confidence Interval

Slide 40 illustrates the distinction between Range, SD, SE, and CI as sample size changes.

Complete the calculation of error metrics for behavioral reaction times (`timing$rt_ms`):

```r
# --- TASK: COMPLETE THE METRICS ---
calc_error_metrics <- function(x, alpha = 0.05) {
  n <- length(x)
  m <- mean(x)
  s <- sd(x)
  
  # Standard Error of the Mean (SE)
  se <- s / sqrt(n)
  
  # Student's t critical value
  t_crit <- qt(1 - alpha/2, df = n - 1)
  
  # 95% Confidence Interval bounds
  ci_lower <- m - t_crit * se
  ci_upper <- m + t_crit * se
  
  return(data.frame(
    Mean = m,
    SD = s,
    SE = se,
    CI_Width = ci_upper - ci_lower
  ))
}

# Compare a small pilot sample (n = 5) vs the full cohort (n = 120)
set.seed(123)
pilot_sample <- sample(timing$rt_ms, size = 5)
full_sample  <- timing$rt_ms

rbind(
  Pilot_N5  = calc_error_metrics(pilot_sample),
  Cohort_N120 = calc_error_metrics(full_sample)
)
```

Notice that as $N$ increases:
* $\text{SD}$ stays approximately constant (it measures population spread).
* $\text{SE}$ and the $\text{CI}$ shrink proportionally to $1/\sqrt{n}$ (they measure precision of the mean estimate).

---

## Step 3: Peer Code Review & Diagnostic Discussion (20 min)

Form pairs and compare outputs:

1. **The Small-$N$ Student's $t$ Penalty:**  
   Compare $t_{n-1, 0.975}$ when $n = 4$ ($df = 3 \Rightarrow t \approx 3.18$) vs. $z_{0.975} = 1.96$. What happens when researchers with small sample sizes use $z$ instead of $t$?
2. **Q-Q Plot Diagnosis:**  
   Inspect the Q-Q plot for `timing$evoked_lat_ms`. What shape appears if you take the natural log (`log(timing$evoked_lat_ms)`) before running `qqnorm()`?