---
title: "Lecture 4: Correlation, Linear Models & Multiplicity"
subtitle: "SISSA PhD in Neuroscience - Practical Lab"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
filters:
  - shinylive
---

## Step 1: Intuition Playground (30 min)

### Leverage, Residuals & Influential Observations
In Ordinary Least Squares (OLS) regression:
$$\hat{Y} = \beta_0 + \beta_1 X + \epsilon, \quad \epsilon \sim \mathcal{N}(0, \sigma^2)$$

Because the regression line minimizes the sum of squared errors $\sum e_i^2$, points located far along the predictor axis ($X$) exert high **leverage**. If their response ($Y$) deviates from the regression trend, they tilt the line disproportionately.

Use the interactive simulator below:
1. Drag the **Outlier Position ($X$)** slider from the center ($X = 0$) to the extreme periphery ($X = 5$).
2. Adjust the **Outlier Vertical Shift ($Y$)** to introduce a large discrepancy.
3. Observe how Cook's distance and the estimated slope $\hat{\beta}_1$ change when an outlier moves from low leverage to high leverage.

```{shinylive-r}
#| standalone: true
#| viewerHeight: 580

library(shiny)

ui <- fluidPage(
  titlePanel("Leverage & Cook's Distance in Regression"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("outlier_x", "Outlier Position (Predictor X):", min = -4, max = 5, value = 4.2, step = 0.2),
      sliderInput("outlier_y", "Outlier Value (Response Y):", min = -10, max = 15, value = -6, step = 0.5),
      helpText("The blue dashed line is the true underlying relationship (slope = 1.2). The red solid line is the fitted OLS model including the outlier.")
    ),
    mainPanel(
      plotOutput("regPlot", height = "440px"),
      verbatimTextOutput("regStats")
    )
  )
)

server <- function(input, output) {
  sim_pts <- reactive({
    set.seed(42)
    x <- rnorm(35, mean = 0, sd = 1.2)
    y <- 1.2 * x + rnorm(35, sd = 0.9)
    
    # Append the interactive outlier
    x_all <- c(x, input$outlier_x)
    y_all <- c(y, input$outlier_y)
    
    mod <- lm(y_all ~ x_all)
    cooks <- cooks.distance(mod)
    
    list(x = x_all, y = y_all, mod = mod, cooks = cooks)
  })

  output$regPlot <- renderPlot({
    res <- sim_pts()
    outlier_idx <- length(res$x)
    is_high_cook <- res$cooks[outlier_idx] > 0.5
    
    par(mar = c(4.5, 4.5, 3, 1))
    plot(res$x[-outlier_idx], res$y[-outlier_idx], pch = 19, col = "steelblue",
         xlim = c(-5, 6), ylim = c(-12, 16),
         xlab = "Predictor (X)", ylab = "Response (Y)",
         main = "Impact of a Single Point on Ordinary Least Squares")
    
    # Highlight the outlier
    points(res$x[outlier_idx], res$y[outlier_idx], pch = 21, bg = "red", cex = 1.8, lwd = 2)
    text(res$x[outlier_idx], res$y[outlier_idx] + 1.2, labels = "Outlier", col = "red", font = 2)
    
    # Ground truth vs Fitted line
    abline(a = 0, b = 1.2, col = "blue", lty = 2, lwd = 2)
    abline(res$mod, col = "firebrick", lwd = 2.5)
    
    legend("topleft", legend = c("True Model (slope = 1.2)", "Fitted OLS Model", "Data Points"),
           col = c("blue", "firebrick", "steelblue"), lty = c(2, 1, NA), pch = c(NA, NA, 19), lwd = 2, bty = "n")
  })
  
  output$regStats <- renderPrint({
    res <- sim_pts()
    outlier_idx <- length(res$x)
    cat("Fitted Slope (Beta 1):", round(coef(res$mod)[2], 3), "(Ground Truth = 1.200)\n")
    cat("Outlier Cook's Distance:", round(res$cooks[outlier_idx], 3), "\n")
    if (res$cooks[outlier_idx] > 0.5) {
      cat("WARNING: Point qualifies as highly influential (Cook's D > 0.5)!\n")
    }
  })
}

shinyApp(ui, server)
```

---

## Step 2: "Code-Along & Fix-It" Lab (85 min)

We use behavioral tracking and neural activity recordings: `data/lecture4_behavior_regression.csv`.

### Exercise 1: Pearson vs. Spearman Correlation

Pearson's $r$ evaluates *linear* associations, whereas Spearman's rank correlation $\rho$ assesses *monotonic* associations.

```{r}
df <- read.csv("../data/lecture4_behavior_regression.csv")

# Inspect relationship between Pupil Diameter and Sensory Firing Rate
plot(df$pupil_diam, df$firing_rate, pch = 19, col = "darkorange",
     xlab = "Normalized Pupil Diameter", ylab = "Firing Rate (Hz)",
     main = "Arousal vs. Firing Rate (Yerkes-Dodson Curve)")
```

Calculate Pearson's and Spearman's correlation coefficients:

```r
# Linear correlation
r_pearson  <- cor(df$pupil_diam, df$firing_rate, method = "pearson")

# Rank correlation
rho_spear  <- cor(df$pupil_diam, df$firing_rate, method = "spearman")

cat("Pearson's r:  ", round(r_pearson, 3), "\n")
cat("Spearman's rho:", round(rho_spear, 3), "\n")
```

**Discussion:** Both coefficients return values close to zero. Why? Neither metric detects non-monotonic non-linear curves (e.g., an inverted-U relationship).

---

### Exercise 2: Fitting Linear Models and Residual Diagnostics

Model hippocampal theta rhythm oscillation frequency as a function of running velocity:

$$\text{theta\_hz}_i = \beta_0 + \beta_1 \cdot \text{speed\_cms}_i + \epsilon_i$$

```r
# --- TASK: FIT OLS REGRESSION ---
fit_theta <- lm(theta_hz ~ speed_cms, data = df)
summary(fit_theta)
```

#### Diagnostic Plots: Checking OLS Assumptions
Never report a linear model without inspecting residuals:
1. **Residuals vs. Fitted:** Assesses linearity and homoscedasticity (constant variance).
2. **Normal Q-Q Plot:** Checks normality of residuals.
3. **Residuals vs. Leverage:** Identifies influential observations via Cook's distance lines.

```r
par(mfrow = c(2, 2))
plot(fit_theta)
```

Look at observation **#12**:
* It sits at extreme leverage ($X \approx 48$) with a large negative residual.
* Cook's distance exceeds typical thresholds ($> 0.5$ or $4/n$).

#### The Fix (Sensitivity Analysis):
Re-fit the model excluding the artifact point and compare slope estimates:

```r
# Exclude observation 12
fit_clean <- lm(theta_hz ~ speed_cms, data = df[-12, ])

cat("Slope WITH outlier:   ", round(coef(fit_theta)[2], 4), "\n")
cat("Slope WITHOUT outlier:", round(coef(fit_clean)[2], 4), "\n")
```

---

### Exercise 3: Multiple Comparisons Correction (Bonferroni vs. FDR)

Our recording rig captured 100 simultaneous local field potential (LFP) channels (`lfp_ch_1` to `lfp_ch_100`). We want to discover which channels correlate significantly with behavioral performance (`behavior_score`).

Run 100 uncorrected hypothesis tests:

```{r}
ch_cols <- paste0("lfp_ch_", 1:100)
p_values <- numeric(100)

for (i in 1:100) {
  test <- cor.test(df[, ch_cols[i]], df$behavior_score)
  p_values[i] <- test$p.value
}

# Number of uncorrected "discoveries" at alpha = 0.05
sum(p_values < 0.05)
```

Without correction, approximately $5\%$ of truly noise channels return false positives by pure chance ($100 \times 0.05 \approx 5$).

Apply **Bonferroni (FWER)** and **Benjamini-Hochberg (FDR)** corrections:

```r
# --- TASK: APPLY MULTIPLE TESTING CORRECTIONS ---

# 1. Family-Wise Error Rate (Bonferroni: alpha / m)
p_bonf <- p.adjust(p_values, method = "bonferroni")

# 2. False Discovery Rate (Benjamini-Hochberg)
p_fdr  <- p.adjust(p_values, method = "BH")

cat("Significant channels (Uncorrected):", sum(p_values < 0.05), "\n")
cat("Significant channels (Bonferroni):  ", sum(p_bonf < 0.05), "\n")
cat("Significant channels (BH / FDR):    ", sum(p_fdr < 0.05), "\n")

# Which exact channels passed FDR threshold?
which(p_fdr < 0.05)
```

---

## Step 3: Peer Code Review & Diagnostic Discussion (20 min)

Form pairs and compare outputs:

1. **Family-Wise Error Rate vs. False Discovery Rate:**  
   When would you recommend using Bonferroni over FDR in neuroscience? (e.g., confirmatory testing of 3 clinical candidate drugs vs. exploratory screening of 20,000 RNA-seq transcripts).
2. **Transformations vs. Polynomials:**  
   Return to Exercise 1 (Pupil vs. Firing Rate). Fit a quadratic term using `lm(firing_rate ~ pupil_diam + I(pupil_diam^2), data = df)`. Inspect `summary()` and the residual plot. Does adding the second-order term resolve the non-linearity?